---
title: "Melbourne Business Establishments: Descriptive Analytics"
author: "Group 10"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  github_document:
    toc: true
    html_preview: false
    fig_width: 10
    fig_height: 6
    df_print: kable
---

```{r setup, include=FALSE}
# Keep the rendered report readable and use consistent figure settings.
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "95%"
)

library(tidyverse)
library(scales)
library(knitr)

# Main comparison years used throughout this report.
YEAR_PRE_COVID <- 2019
YEAR_TROUGH <- 2022
YEAR_LATEST <- 2024

# South Yarra has a very small 2019 baseline (54 establishments). Percentage
# changes from such a small base can dominate a ranking, so rankings require
# at least 100 establishments in the baseline year.
MIN_BASELINE_FOR_RANKING <- 100

# Shared colours keep all figures visually consistent.
COL_DARK <- "#2C3E50"
COL_BLUE <- "#4C78A8"
COL_LIGHT_BLUE <- "#A8BDD0"
COL_NEGATIVE <- "#D95F59"
COL_POSITIVE <- "#2A9D8F"
COL_NEUTRAL <- "#D9E2E8"

# Return percentage change while protecting against division by zero.
pct_change <- function(new_value, old_value) {
  if_else(old_value == 0, NA_real_, 100 * (new_value - old_value) / old_value)
}
```

# 1. Research question and scope

**Research question:** How have small and medium-sized business establishment
patterns evolved across Melbourne precincts and industries from 2002 to 2024?

This report describes establishment patterns across the full period. It uses
2019 as the pre-COVID baseline, 2022 as the observed low point and 2024 as the
latest comparison. It does not treat every change as a COVID-19 effect.

The analysis follows three rules:

1. A CLUE establishment is a business location, not necessarily a unique legal
   enterprise.
2. CLUE uses a two-year rolling survey cycle. The full annual series is shown
   for context, but substantive comparisons use years at least two years apart.
3. Establishment count is the primary measure. Some job values are suppressed
   for confidentiality and are not imputed in this report.

# 2. Import the two CLUE files

Place the supplied CSV files in the repository's `data/` folder. The fallback
paths also allow the report to run from the original project folder.

```{r import-data}
# Look for the data in the standard repository data folder first, then fall
# back to the current or parent folder. This keeps the report portable when it
# is stored inside analysis/ and avoids computer-specific absolute paths.
data_candidates <- c("data", "../data", ".", "..")

DATA_DIR <- data_candidates[
  file.exists(file.path(
    data_candidates,
    "business-establishments-and-jobs-data-by-business-size-and-anzsic.csv"
  ))
][1]

if (is.na(DATA_DIR)) {
  stop("Data files not found. Place both supplied CSV files in the data/ folder.")
}

location_path <- file.path(
  DATA_DIR,
  "business-establishments-with-address-and-industry-classification.csv"
)

aggregate_path <- file.path(
  DATA_DIR,
  "business-establishments-and-jobs-data-by-business-size-and-anzsic.csv"
)

# Stop with a clear error if either required dataset is missing.
stopifnot(file.exists(location_path), file.exists(aggregate_path))

location_raw <- read_csv(location_path, show_col_types = FALSE)
aggregate_raw <- read_csv(aggregate_path, show_col_types = FALSE)
```

## 2.1 File overview

```{r file-overview}
file_overview <- tibble(
  dataset = c("Location-level file", "Aggregated file"),
  rows = c(nrow(location_raw), nrow(aggregate_raw)),
  columns = c(ncol(location_raw), ncol(aggregate_raw))
)

kable(file_overview, format.args = list(big.mark = ","))
```

# 3. Clean and validate the data

## 3.1 Aggregated file

The aggregated file is the main source for establishment trends because it
contains consistent totals by year, precinct, industry and business size.

```{r clean-aggregate}
aggregate <- aggregate_raw %>%
  rename(
    census_year = `Census year`,
    clue_small_area = `CLUE small area`,
    anzsic_industry = `ANZSIC indusrty`,
    business_size = `Business size`,
    total_establishments = `Total establishments`,
    total_jobs = `Total jobs`
  ) %>%
  mutate(
    census_year = as.integer(census_year),
    clue_small_area = as.factor(clue_small_area),
    anzsic_industry = as.factor(anzsic_industry),
    business_size = factor(
      business_size,
      levels = c(
        "Non employing",
        "Small business",
        "Medium business",
        "Large business"
      )
    ),
    total_establishments = as.numeric(total_establishments),
    total_jobs = as.numeric(total_jobs),
    jobs_suppressed = is.na(total_jobs)
  )

# These checks detect accidental changes to the supplied dataset or import.
stopifnot(
  nrow(aggregate) == 13353,
  n_distinct(aggregate$census_year) == 23,
  n_distinct(aggregate$clue_small_area) == 13,
  sum(is.na(aggregate$total_establishments)) == 0,
  sum(aggregate$jobs_suppressed) == 4252,
  sum(duplicated(aggregate)) == 0
)
```

## 3.2 Location-level file

The location-level file provides addresses, detailed ANZSIC classifications and
coordinates. It also contains records coded as `Vacant Space`. These records are
excluded from business counts because they represent empty locations rather
than operating establishments.

```{r clean-location}
location <- location_raw %>%
  mutate(
    census_year = as.integer(census_year),
    is_vacant_space =
      industry_anzsic4_code == 0 &
      industry_anzsic4_description == "Vacant Space",
    is_operating = industry_anzsic4_code > 0
  )

location_operating <- location %>%
  filter(is_operating)

location_quality_check <- location %>%
  summarise(
    all_location_records = n(),
    vacant_space_records = sum(is_vacant_space),
    operating_records = sum(is_operating)
  )

kable(location_quality_check, format.args = list(big.mark = ","))

stopifnot(
  sum(location$is_vacant_space) == 68299,
  all(location$is_vacant_space == !location$is_operating),
  sum(duplicated(location)) == 0
)
```

The 68,299 Vacant Space observations are records across all 23 census years.
They are not 68,299 unique buildings and are not a current vacancy count.

## 3.3 Reconcile the two files

```{r reconcile-files}
location_yearly <- location %>%
  group_by(census_year) %>%
  summarise(
    operating_location_records = sum(is_operating),
    .groups = "drop"
  )

aggregate_yearly <- aggregate %>%
  group_by(census_year) %>%
  summarise(
    official_establishments = sum(total_establishments),
    .groups = "drop"
  )

reconciliation <- aggregate_yearly %>%
  left_join(location_yearly, by = "census_year") %>%
  mutate(
    difference = operating_location_records - official_establishments
  )

kable(
  reconciliation %>%
    filter(census_year %in% c(YEAR_PRE_COVID, YEAR_TROUGH, YEAR_LATEST)),
  format.args = list(big.mark = ",")
)

# The two sources closely align after Vacant Space is removed. The small
# remaining difference never exceeds 25 records in any census year.
stopifnot(max(abs(reconciliation$difference)) <= 25)
```

# 4. Variables of interest

## 4.1 Numerical variables

```{r numerical-variables}
numerical_variables <- tribble(
  ~variable, ~role, ~use_in_this_report,
  "census_year", "Time variable", "Long-run trend and comparison-year selection",
  "total_establishments", "Primary outcome", "Business establishment counts",
  "total_jobs", "Secondary outcome", "Reported only; missing values are not imputed",
  "longitude / latitude", "Spatial coordinates", "Reserved for later spatial analysis"
)

kable(numerical_variables)
```

## 4.2 Categorical variables

```{r categorical-variables}
categorical_variables <- tribble(
  ~variable, ~levels, ~use_in_this_report,
  "clue_small_area", n_distinct(aggregate$clue_small_area), "Precinct comparisons",
  "anzsic_industry", n_distinct(aggregate$anzsic_industry), "Industry comparisons",
  "business_size", n_distinct(aggregate$business_size), "Business-size comparisons",
  "industry_anzsic4_description",
  n_distinct(location_operating$industry_anzsic4_description),
  "Detailed industry checks in later analysis"
)

kable(categorical_variables)
```

# 5. Summary statistics for key numerical variables

One row in the aggregated file is a year × precinct × industry × business-size
cell. The following statistics describe these cells, not individual businesses.

```{r overall-summary-statistics}
summary_statistics <- aggregate %>%
  summarise(
    cells = n(),
    establishment_mean = mean(total_establishments),
    establishment_median = median(total_establishments),
    establishment_sd = sd(total_establishments),
    establishment_min = min(total_establishments),
    establishment_max = max(total_establishments),
    jobs_reported_cells = sum(!jobs_suppressed),
    jobs_missing_cells = sum(jobs_suppressed),
    reported_jobs_mean = mean(total_jobs, na.rm = TRUE),
    reported_jobs_median = median(total_jobs, na.rm = TRUE)
  )

kable(
  summary_statistics,
  digits = 1,
  format.args = list(big.mark = ",")
)
```

The job summaries describe reported values only. The 4,252 blank job cells are
confidentiality suppressions associated with cells containing one or two
establishments; they are retained and not replaced with zero.

## 5.1 Distribution of establishment counts across aggregated cells

```{r establishment-histogram, fig.cap="Distribution of establishment counts across aggregated cells."}
ggplot(aggregate, aes(total_establishments)) +
  geom_histogram(
    bins = 40,
    fill = COL_BLUE,
    colour = "white",
    linewidth = 0.2
  ) +
  scale_x_log10(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Most aggregated cells contain a small number of establishments",
    subtitle = "The horizontal axis uses a log scale to show the long right tail",
    x = "Establishments in each aggregated cell (log scale)",
    y = "Number of cells"
  ) +
  theme_minimal(base_size = 12)
```

## 5.2 Distribution by business size

```{r establishment-boxplot, fig.cap="Cell-level establishment counts by business-size category."}
ggplot(aggregate, aes(business_size, total_establishments, fill = business_size)) +
  geom_boxplot(outlier.alpha = 0.15, show.legend = FALSE) +
  scale_y_log10(labels = comma) +
  labs(
    title = "Cell counts vary widely within each business-size group",
    subtitle = "The vertical axis uses a log scale",
    x = NULL,
    y = "Establishments in each aggregated cell (log scale)"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

# 6. Grouped descriptive statistics

## 6.1 City-wide totals by comparison year

```{r city-year-summary}
city_year_summary <- aggregate %>%
  filter(census_year %in% c(YEAR_PRE_COVID, YEAR_TROUGH, YEAR_LATEST)) %>%
  group_by(census_year) %>%
  summarise(
    establishments = sum(total_establishments),
    reported_jobs = sum(total_jobs, na.rm = TRUE),
    suppressed_job_cells = sum(jobs_suppressed),
    .groups = "drop"
  )

kable(city_year_summary, format.args = list(big.mark = ","))
```

## 6.2 Precinct summary for 2019 and 2024

```{r precinct-summary}
precinct_summary <- aggregate %>%
  filter(census_year %in% c(YEAR_PRE_COVID, YEAR_LATEST)) %>%
  group_by(census_year, clue_small_area) %>%
  summarise(
    establishments = sum(total_establishments),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = census_year,
    values_from = establishments,
    names_prefix = "est_"
  ) %>%
  mutate(
    change_count = est_2024 - est_2019,
    change_pct = pct_change(est_2024, est_2019),
    eligible_for_pct_ranking = est_2019 >= MIN_BASELINE_FOR_RANKING
  ) %>%
  arrange(change_pct)

kable(
  precinct_summary,
  digits = 1,
  format.args = list(big.mark = ",")
)
```

## 6.3 Business-size summary for 2019 and 2024

```{r business-size-summary}
size_summary <- aggregate %>%
  filter(census_year %in% c(YEAR_PRE_COVID, YEAR_LATEST)) %>%
  group_by(census_year, business_size) %>%
  summarise(
    establishments = sum(total_establishments),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = census_year,
    values_from = establishments,
    names_prefix = "est_"
  ) %>%
  mutate(
    change_count = est_2024 - est_2019,
    change_pct = pct_change(est_2024, est_2019)
  )

kable(
  size_summary,
  digits = 1,
  format.args = list(big.mark = ",")
)
```

## 6.4 Industry summary for 2019 and 2024

```{r industry-summary}
industry_summary <- aggregate %>%
  filter(census_year %in% c(YEAR_PRE_COVID, YEAR_LATEST)) %>%
  group_by(census_year, anzsic_industry) %>%
  summarise(
    establishments = sum(total_establishments),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = census_year,
    values_from = establishments,
    names_prefix = "est_"
  ) %>%
  mutate(
    change_count = est_2024 - est_2019,
    change_pct = pct_change(est_2024, est_2019)
  ) %>%
  arrange(change_count)

kable(
  industry_summary,
  digits = 1,
  format.args = list(big.mark = ",")
)
```

# 7. Visual analysis

## 7.1 Long-run city-wide establishment trend

The annual points show the published series. They are used to describe the
long-run shape, not to estimate year-on-year entry or exit.

```{r city-trend-plot, fig.cap="City-wide establishment trend, 2002–2024."}
city_trend <- aggregate %>%
  group_by(census_year) %>%
  summarise(
    establishments = sum(total_establishments),
    .groups = "drop"
  )

ggplot(city_trend, aes(census_year, establishments)) +
  geom_line(linewidth = 1, colour = COL_DARK) +
  geom_point(size = 2, colour = COL_DARK) +
  geom_vline(
    xintercept = 2020,
    linetype = "dashed",
    colour = COL_NEGATIVE
  ) +
  annotate(
    "text",
    x = 2020.3,
    y = max(city_trend$establishments),
    label = "COVID-19",
    hjust = 0,
    colour = COL_NEGATIVE,
    size = 3.5
  ) +
  scale_x_continuous(breaks = seq(2002, 2024, 2)) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Melbourne had fewer establishments after 2019",
    subtitle = "Comparisons use years at least two years apart",
    x = "Census year",
    y = "Business establishments"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 7.2 Precinct baseline before COVID-19

```{r precinct-baseline-plot, fig.cap="Precinct establishment counts in 2019."}
precinct_baseline <- aggregate %>%
  filter(census_year == YEAR_PRE_COVID) %>%
  group_by(clue_small_area) %>%
  summarise(
    establishments = sum(total_establishments),
    .groups = "drop"
  ) %>%
  mutate(
    city_share = establishments / sum(establishments),
    emphasis = if_else(
      clue_small_area == "Melbourne (CBD)",
      "CBD",
      "Other precincts"
    ),
    label = paste0(comma(establishments), " (", percent(city_share, 0.1), ")")
  )

ggplot(
  precinct_baseline,
  aes(
    establishments,
    fct_reorder(clue_small_area, establishments),
    fill = emphasis
  )
) +
  geom_col(width = 0.68) +
  geom_text(aes(label = label), hjust = -0.05, size = 3.5) +
  scale_fill_manual(
    values = c("CBD" = COL_DARK, "Other precincts" = COL_LIGHT_BLUE),
    guide = "none"
  ) +
  scale_x_continuous(labels = comma, limits = c(0, 10800)) +
  labs(
    title = "Precincts entered COVID-19 from very different baselines",
    subtitle = "The CBD contained 56.5% of establishments in 2019",
    x = "Business establishments in 2019",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

## 7.3 Precinct recovery gap, 2019 to 2024

South Yarra is shown in the summary table but excluded from this percentage
ranking because its 2019 baseline was only 54 establishments.

```{r precinct-change-plot, fig.cap="Percentage change in establishments by precinct, excluding baselines below 100."}
precinct_rank <- precinct_summary %>%
  filter(eligible_for_pct_ranking)

ggplot(
  precinct_rank,
  aes(
    change_pct,
    fct_reorder(clue_small_area, change_pct)
  )
) +
  geom_col(fill = COL_NEGATIVE, width = 0.68) +
  geom_vline(xintercept = 0, colour = "grey40") +
  geom_text(
    aes(
      x = -0.3,
      label = percent(change_pct / 100, accuracy = 0.1)
    ),
    hjust = 1,
    colour = "white",
    size = 3.4
  ) +
  labs(
    title = "No precinct returned to its 2019 level by 2024",
    subtitle = "The chart excludes South Yarra from percentage ranking because its baseline was only 54",
    x = "Change in establishments, 2019 to 2024",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

## 7.4 Change by business size

```{r business-size-change-plot, fig.cap="Change in establishment counts by business size, 2019–2024."}
ggplot(
  size_summary,
  aes(change_count, fct_reorder(business_size, change_count))
) +
  geom_col(
    aes(fill = change_count >= 0),
    width = 0.65,
    show.legend = FALSE
  ) +
  geom_vline(xintercept = 0, colour = "grey40") +
  geom_text(
    aes(
      label = comma(change_count),
      hjust = if_else(change_count < 0, 1.1, -0.1)
    ),
    colour = COL_DARK,
    size = 3.7
  ) +
  scale_fill_manual(
    values = c("TRUE" = COL_POSITIVE, "FALSE" = COL_NEGATIVE)
  ) +
  scale_x_continuous(
    labels = comma,
    limits = c(-2700, 300)
  ) +
  labs(
    title = "The establishment decline was concentrated in Small businesses",
    subtitle = "Small businesses lost 2,434 establishments from 2019 to 2024",
    x = "Change in establishments, 2019 to 2024",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

## 7.5 Industries with the largest numerical changes

```{r industry-change-plot, fig.cap="Industries with the largest absolute establishment changes, 2019–2024."}
industry_plot_data <- industry_summary %>%
  slice_max(abs(change_count), n = 10, with_ties = FALSE)

ggplot(
  industry_plot_data,
  aes(change_count, fct_reorder(anzsic_industry, change_count))
) +
  geom_col(
    aes(fill = change_count >= 0),
    width = 0.68,
    show.legend = FALSE
  ) +
  geom_vline(xintercept = 0, colour = "grey40") +
  scale_fill_manual(
    values = c("TRUE" = COL_POSITIVE, "FALSE" = COL_NEGATIVE)
  ) +
  scale_x_continuous(labels = comma) +
  labs(
    title = "Professional services and retail recorded the largest losses",
    subtitle = "Ten industries with the largest numerical changes, 2019 to 2024",
    x = "Change in establishments, 2019 to 2024",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank())
```

# 8. Main findings

```{r headline-findings}
headline_findings <- tibble(
  finding = c(
    "City-wide establishments fell from 16,713 in 2019 to 14,338 in 2024, a 14.2% decline.",
    "None of the 13 precincts returned to its 2019 establishment level by 2024.",
    "Small businesses lost 2,434 establishments, more than the total city-wide net decline.",
    "Large businesses added 47 establishments, partly offsetting the small-business decline.",
    "Professional, Scientific and Technical Services recorded the largest numerical loss (-638), followed by Retail Trade (-393)."
  )
)

kable(headline_findings, col.names = "Finding")
```

# 9. Limitations and next step

- The analysis identifies patterns but does not establish causal effects of
  COVID-19.
- Job values are incomplete because CLUE suppresses cells with one or two
  establishments. This report does not impute them.
- Percentage rankings exclude precincts with fewer than 100 establishments in
  2019, although those precincts remain in the underlying tables.
- The next stage will compare precinct × industry × business-size patterns at
  multi-year intervals rather than calculate year-on-year entry or exit rates.

# 10. Reproducibility information

```{r session-information}
sessionInfo()
```
